---
title: "Cyclistic Data - R Notebook Demo"
author: Aucoeur Ngo
date: June 26, 2024
output: 
  github_document:
    toc: true
    toc_depth: 2
---

# Case Study: How Does a Bike-Share Navigate Speedy Success?

Coursera Course 8 Google Data Analytics <https://www.coursera.org/learn/google-data-analytics-capstone/>

## ASK

### Business Task:

Analyze historical bike data to understand how annual members and casual riders use Cyclistic bike differently to make recommendations for converting casual riders into annual members.

## PREPARE

We are using 12 months of data provided by **Motivate International Inc.** under this license: <https://divvybikes.com/data-license-agreement>

### Dataset Details

```{r}
# Install if necessary
install.packages("tidyverse")
install.packages("lubridate")
install.packages("ggplot2")
install.packages("dplyr")
install.packages("readr")
```

```{r}
# Load libraries necessary for all steps in the analysis proccess of Cyclistic's historical data
library(tidyverse)  
library(lubridate)  
library(ggplot2)  
library(dplyr)
library(readr)
```

Import dataset

```{r}
# Import Dataset
# CyclisticData <- read_csv("Trip Data/2024-2023-12mo-tripdata.zip")

# Import Data by month June 2023 - May 2024
X202405_divvy_tripdata <- read_csv("Trip Data/202405-divvy-tripdata.csv")
X202404_divvy_tripdata <- read_csv("Trip Data/202404-divvy-tripdata.csv")
X202403_divvy_tripdata <- read_csv("Trip Data/202403-divvy-tripdata.csv")
X202402_divvy_tripdata <- read_csv("Trip Data/202402-divvy-tripdata.csv")
X202401_divvy_tripdata <- read_csv("Trip Data/202401-divvy-tripdata.csv")
X202312_divvy_tripdata <- read_csv("Trip Data/202312-divvy-tripdata.csv")
X202311_divvy_tripdata <- read_csv("Trip Data/202311-divvy-tripdata.csv")
X202310_divvy_tripdata <- read_csv("Trip Data/202310-divvy-tripdata.csv")
X202309_divvy_tripdata <- read_csv("Trip Data/202309-divvy-tripdata.csv")
X202308_divvy_tripdata <- read_csv("Trip Data/202308-divvy-tripdata.csv")
X202307_divvy_tripdata <- read_csv("Trip Data/202307-divvy-tripdata.csv")
X202306_divvy_tripdata <- read_csv("Trip Data/202306-divvy-tripdata.csv")

```

```{r}
# Merge monthly datasets into one
Tripdata_12mo <- rbind(X202306_divvy_tripdata, X202307_divvy_tripdata, X202308_divvy_tripdata, X202309_divvy_tripdata, X202310_divvy_tripdata, X202311_divvy_tripdata, X202312_divvy_tripdata, X202401_divvy_tripdata, X202402_divvy_tripdata, X202403_divvy_tripdata, X202404_divvy_tripdata, X202405_divvy_tripdata)

head(Tripdata_12mo)
# glimpse(Tripdata_12mo)
summary(Tripdata_12mo)
```

```{r}
glimpse(Tripdata_12mo)
# Show dimensions of the table
dim(Tripdata_12mo)

# Show column names
colnames(Tripdata_12mo)
```

```{r}
# Other summary / dataset info functions
# summary(CyclisticData)
# as_tibble(CyclisticData)
```

Cleaning dataset
```{r}
install.packages("here")
install.packages("skimr")
install.packages("janitor")
```

```{r}
library(here)
library(skimr)
library(janitor)
```

```{r}


# Removing NAs / null from data
CleanedData <- Tripdata_12mo %>% na.omit()
CleanedDataNoDupes <- CleanedData %>% distinct(ride_id, .keep_all = TRUE)

print(paste("Removed", nrow(CleanedData) - nrow(CleanedDataNoDupes), "duplicates with same ride_id"))

glimpse(CleanedData)
dim(CleanedData)
# as_tibble(CleanedData)
```


```{r}
# Add columns for Month, weekday
# Calculating ride_length from started_at & ended_at
TransformedCleaned <- CleanedData %>%
  mutate(ride_length = ended_at - started_at) %>%
  mutate(month = month(started_at, label = TRUE)) %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  #select(ride_id, rideable_type, month, weekday, ride_length, member_casual)
  select(ride_id, rideable_type, month, weekday,started_at, ended_at, ride_length, member_casual, start_station_name, end_station_name, start_lat, start_lng, end_lat, end_lng)

# Adding columns for Country, State, City 
StationData <- 
  cbind(TransformedCleaned, 
        country = "united states",
        state = "illinois",
        city = "chicago") 

head(StationData)
```
### Exporting Dataframes

```{r}
write.csv(TransformedCleaned, "Trip Data/cleaned/merged-monthly-daily.csv")
write.csv(StationData, "Trip Data/cleaned/station_data.csv")
```

If dataframes are still too large, consider:
  - Exporting by month
  - Try `sample()` to export a smaller set https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/sample


```{r}
CountMemberCasual <- TransformedCleaned %>% count(member_casual)
print(CountMemberCasual)
```

## SHARE

### Visualization in R

Monthly Usage
```{r}
# Prepping data for visualization
MonthlyRidesByMemberType <- TransformedCleaned %>%
  group_by(month, member_casual) %>%
  summarize(total_rides = n())

# Column Chart Viz
MonthlyRidesViz <- ggplot(data = MonthlyRidesByMemberType, 
                       mapping = aes(x = month, y = total_rides, fill = member_casual)) +
  geom_col() +
  labs(title = "Bike Usage Frequency", subtitle = "Rides per month", x = "Month", y = "Total Rides")


MonthlyRidesViz
```
Conclusion: Overall usage is greatest from May - Oct with lowest usage corresponding to Dec - Jan (winter months).  We also observe that members have more consistent usage across the 12 months, compared to casual users.

Usage By Day
```{r}
# Create the dataframe for the vis / prepare the data
DailyRidesByMemberType <- TransformedCleaned %>%
    group_by(weekday, member_casual) %>%
  summarize(total_rides = n())

# Generate the viz
DailyRidesViz <- ggplot(data = DailyRidesByMemberType,
                        aes(x =  weekday, y = total_rides, fill = member_casual)) +
                  geom_col(position = "dodge") + 
                  labs(title = "Usage By Day", subtitle = "Total rides per day",
                       x = "Day", y = "Total Number of Rides")

DailyRidesViz

```

## Areas to Branch Off / Dive Deeper

- How does the Daily chart look when only looking at data from May - Oct?
- How do we display the monthly chart in a way that is easy to see at a glance when our data is June 2023 - May 2024 (as opposed to Jan - Dec)
- What other charts and graphs would provide good "at-a-glance" insights?








